Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-14 arxiv/intro.tex
address vaggos comments

Hard parameter sharing is intuitively appealing as an inductive transfer mechanism.
It reduces overfitting by acting as a regularizer \cite{R17}.
For example, by restricting the shared space's size, HPS encourages information sharing among multiple tasks \cite{KD12}.
Another source of inductive bias comes from the tasks and depends on datasets' properties such as sample sizes and task covariances \cite{WZR20}.
However, how these dataset properties impact HPS has not been well established.
%It becomes increasingly important to understand HPS' formal generalization properties.
Part of the challenge may be that HPS' generalization performance depends intricately on the sample size ratios and covariate shifts between tasks, and is not amenable to standard concentration results.
Previous results based on Rademacher complexity or VC dimensions have considered cases when all tasks' sample sizes are equal to logarithm factors of the feature dimension \cite{B00,MPR16}, and when all tasks' sample sizes increase simultaneously \cite{AZ05,M06}.
%For, the generalization error scales down as the sample sizes of all tasks increase, when applied to the multi-task setting \cite{B00,AZ05,M06,MPR16,WZR20}.

\begin{figure*}[!t]
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\includegraphics[width=0.45\textwidth,valign=t]{figures/mtl_model_arch.pdf}
		\caption{A hard parameter sharing architecture}
	\end{subfigure}\hfill
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\includegraphics[width=0.6417\textwidth,valign=t]{figures/sample_ratio_c2_400.eps}
		\caption{Varying sample size ratio}
		\label{fig_intro_sample_size_b}
	\end{subfigure}
	\vspace{-0.1in}
	\caption{An illustrative example of our result:
	Consider the prediction loss of hard parameter sharing (left) for task two, given two linear regression tasks.
	Increasing task one's sample size decreases task two's prediction loss initially, but increases afterwards. This phenomenon occurs due to different bias-variance tradeoffs as the sample size ratio increases. Our result provides an estimated loss (solid line) that accurately matches the empirical loss (dotted line).
	See Section \ref{sec_simulation} for the precise setting.}
	\label{fig_intro_sample_size}
	\vspace{-0.1in}
\end{figure*}

This paper presents new techniques to study hard parameter sharing and establish a number of new results.
We consider regression analysis, which is arguably one of the most fundamental problems in statistics and machine learning.
We are interested in the \textit{high-dimensional} setting, where each dataset's sample size and feature dimension grow linearly at a fixed ratio.
This is motivated by many multi-task learning applications, where the amount of labeled data from each dataset is usually insufficient for learning a single task.
For example, this is the case if a dataset's sample size is only a small constant factor of the feature dimension.
The high-dimensional setting is challenging but is crucial for understanding how datasets' sample sizes impact generalization performance.

>>>>

Hard parameter sharing is an intuitively appealing mechanism for inductive transfer, and reduces overfitting by acting as a regularizer \cite{R17}.
For example, by restricting the shared space's size, HPS encourages information sharing among multiple tasks \cite{KD12}.
Another source of inductive bias comes from the tasks and depends on datasets' properties such as sample sizes and task covariances \cite{WZR20}.
However, how these dataset properties impact HPS has not been well established.
%It becomes increasingly important to understand HPS' formal generalization properties.
Part of the challenge may be that HPS' generalization performance depends intricately on the sample size ratios and covariate shifts between tasks, and is not amenable to standard concentration results.
Previous results based on Rademacher complexity or VC dimensions have considered cases when all tasks' sample sizes are equal to logarithmic factors of the feature dimension \cite{B00,MPR16}, and when all tasks' sample sizes increase simultaneously \cite{AZ05,M06}.
%For, the generalization error scales down as the sample sizes of all tasks increase, when applied to the multi-task setting \cite{B00,AZ05,M06,MPR16,WZR20}.

\begin{figure*}[!t]
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\includegraphics[width=0.45\textwidth,valign=t]{figures/mtl_model_arch.pdf}
		\caption{A hard parameter sharing architecture}
		\label{fig_intro_arch}
	\end{subfigure}\hfill
	\begin{subfigure}[t]{0.5\textwidth}
		\centering
		\includegraphics[width=0.745\textwidth,valign=t]{figures/sample_ratio_c2_400.eps}
		\caption{Varying sample size ratio}
		\label{fig_intro_sample_size_b}
	\end{subfigure}
	\vspace{-0.1in}
	\caption{
	Left: an illustrative picture of HPS.
	Right: an illustrative example of using HPS for two tasks $X_1, Y_1$ and $X_2, Y_2$ with sample size $n_1, n_2$, respectively.
	Increasing $n_1/n_2$ decreases task two's prediction loss initially, but increases afterwards. This phenomenon occurs due to different bias-variance tradeoffs as $n_1/n_2$ increases. Our result provides an estimated loss (solid line) that accurately matches the empirical loss (dotted line).
	See Section \ref{sec_simulation} for the precise setting.}
	\label{fig_intro_sample_size}
	\vspace{-0.1in}
\end{figure*}

This paper presents new techniques to study hard parameter sharing and establishes a number of new results.
We consider regression analysis, which is arguably one of the most fundamental problems in statistics and machine learning.
We are interested in the \textit{high-dimensional} setting, where each dataset's sample size and feature dimension grow linearly instead of logarithmicly.
This setting captures the fact that in many applications, a single task's sample size is usually insufficient for accurate learning.
For example, if a dataset's sample size is only a constant factor of dimension in linear regression, the variance is also constant (cf. Fact \ref{fact_tr}).
The high-dimensional setting is challenging but is crucial for understanding how datasets' sample sizes impact generalization performance.

### 2020-10-14 arxiv/setup.tex
address vaggos comments

We assume that there are multiple datasets that all follow (potentially different) linear models.
In each dataset, suppose the feature vector is $x = {\Sigma}^{1/2} z$, where $z \in \real^p$ has i.i.d entries with zero mean and unit variance, and the population covariance matrix $\Sigma \in\real^{p \times p}$ is deterministic and positive semidefinite.

We study a hard parameter sharing architecture to learn jointly from multiple datasets.
In this architecture, there is a shared feature representation layer $B\in\real^{p\times r}$ for all tasks and a separate output layer $A_i \in \real^r$ for every task $i$.
Suppose we have $t$ datasets.
For each dataset $i$ from $1$ to $t$, let $n_i$ denote its sample size, $X^{(i)} \in \real^{n_i \times p}$ denote its feature covariates, and $Y^{(i)} \in \real^{n_i}$ denote all the labels.
%The width of $B$, denoted by $r$, plays an important role in regularization.
%As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
%Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
Let $A = [A_1, A_2, \dots, A_t] \in \real^{r \times t}$.
We study the following optimization objective.
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
%	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
\vspace{-0.1in}
\begin{align}\label{eq_mtl}
			f(A, B) = \sum_{i=1}^t \norm{X^{(i)} B A_i - Y^{(i)}}^2.
\end{align}
%\vspace{-0.1in}
Given a solution $(\hat{A}, \hat{B})$ of $f(A, B)$, let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the HPS estimator for task $i$.
Some critical questions are:
(i) How well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) For datasets with different sample sizes and covariate shifts, how do they affect the estimator?

\textbf{Main results.}
Our first result (Theorem \ref{thm_many_tasks}) applies to the multi-label prediction setting where all datasets have the same features, and we want to make multiple predictions (see e.g. \cite{hsu2009multi}).
We analyze the global minimizer of $f(A, B)$, and provide a sharp generalization bound on its (out-of-sample) prediction loss for any task.
This case is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
We show that the prediction loss of hard parameter sharing admits a clean bias-variance decomposition.
Our results imply that hard parameter sharing helps by reducing the variance compared to single-task learning, but hurts by increasing the bias.
Our second result (Theorem \ref{thm_main_RMT}) applies to two tasks with arbitrarily different sample size ratios and population covariance matrices.
We analyze the local minimizers of $f(A, B)$ and provide a sharp generalization bound on both tasks' prediction loss.
Despite its simplicity, we show several interesting phenomena by varying sample sizes and covariate shifts in this setting.
See Figure \ref{fig_intro_sample_size} for an illustration.

Consequently, using our precise loss estimates, we observe several qualitative properties of hard parameter sharing for varying datasets' properties.
%\begin{enumerate}

\textit{Sample efficiency (Example \ref{ex_same_cov})}:
	One advantage of combining multiple datasets is that the requirement for labeled data reduces compared to STL, a phenomenon that \citet{ZSSGM18} has observed empirically.
	Our results further imply that HPS's sample efficiency depends on model-specific variances across tasks vs. the noise variance, and is generally high when the latter is large.

\textit{Sample size ratio (Example \ref{ex_sample_ratio})}: Increasing one task's sample size does not always help to reduce another task's loss. In a simplified setting, we find that the task loss either decreases first before increasing afterwards, or decreases monotonically depending on how fast the bias increases. These two trends result from different bias-variance tradeoffs.

@@

There are two main ideas in our analysis. The proof of our first result uses a geometric intuition that hard parameter sharing finds a ``rank-$r$'' approximation of the datasets.
We carefully keep track of the concentration error between the global minimum of $f(A, B)$ and its population version.
The proof of our second result is significantly more involved because of different sample sizes and covariate shifts. Using recently developed techniques from random matrix theory \cite{Anisotropic}, we show that the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts converges to a deterministic diagonal matrix asymptotically (cf. Theorem \ref{thm_main_RMT}). % Moreover, we can obtain a sharp bound on the concentration error.
% to obtain a sharp estimate on the..., which is commonly referred to as the \emph{local law}. 
%\HZ{add several sentences on the technical insight} 
One limitation of our analysis is that for two tasks with different sample sizes, there is an extra error term in the prediction loss of hard parameter sharing (cf. equation \eqref{cor_MTL_error}), which can be large for very small $n_1$. This requires studying the singular value spectrum of certain non-symmetric random matrices, which is an intriguing open question (see Section \ref{sec_conclude} for more discussion).
% plus an identity matrix,  due to lack of freeness
%\HZ{add why this is challenging}

Finally, we discuss the practical implications of work.
Our sample size ratio study implies a concrete progressive training procedure that gradually adds more data until performance drops.
For example, in the setting of Figure \ref{fig_intro_sample_size_b}, this procedure will stop right at the minimum of the local basin.
We conduct further studies of this procedure on six text classification datasets and observe that it reduces the computational cost by $65\%$ compared to a standard round-robin training procedure while keeping the average accuracy of all tasks simultaneously.

>>>>

%We assume that there are multiple datasets that all follow (potentially different) linear models.
%In each dataset, suppose the feature vector is $x = {\Sigma}^{1/2} z$, where $z \in \real^p$ has i.i.d entries with zero mean and unit variance, and the population covariance matrix $\Sigma \in\real^{p \times p}$ is deterministic and positive semidefinite.
Suppose we have $t$ datasets.
For each dataset $i$ from $1$ to $t$, let $n_i$ denote its sample size.
Let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates.
We assume that the label vector $Y^{(i)} \in \real^{n_i}$ follows a linear model plus random noise.
We study the standard hard parameter sharing architecture to learn jointly from multiple datasets.
In this architecture, there is a shared feature representation layer $B\in\real^{p\times r}$ for all tasks and a separate output layer $A_i \in \real^r$ for every task $i$.
See Figure \ref{fig_intro_arch} for an illustration.
%The width of $B$, denoted by $r$, plays an important role in regularization.
%As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
%Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
We study the following minimization problem:
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
%	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
%\vspace{-0.1in}
\begin{align}\label{eq_mtl}
			f(A, B) = \sum_{i=1}^t \norm{X^{(i)} B A_i - Y^{(i)}}^2,
\end{align}
%\vspace{-0.1in}
where $A = [A_1, A_2, \dots, A_t] \in \real^{r \times t}$.
Given a solution $(\hat{A}, \hat{B})$ of $f(A, B)$ (which we will specify later), let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the HPS estimator for task $i$.
The critical questions here are:
(i) How well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) For datasets with different sample sizes and covariate shifts, how do they affect the estimator?

\textbf{Main results.}
Our first result (Theorem \ref{thm_many_tasks}) applies to the multi-label prediction setting where all datasets have the same features, and we want to make multiple predictions (see e.g. \cite{hsu2009multi}).
We analyze the global minimizer of $f(A, B)$, and provide a sharp generalization bound on its (out-of-sample) prediction loss for any task.
This case is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
We show that the prediction loss of HPS admits a clean bias-variance decomposition.
Our results imply that in the same covariates setting where the sample sizes are also equal, HPS helps by reducing the variance compared to single-task learning, but hurts by increasing the bias.

Our second result (Theorem \ref{thm_main_RMT}) applies to two tasks with arbitrarily different sample size ratios and population covariance matrices.
We analyze the local minimizers of $f(A, B)$ and provide a sharp generalization bound on both tasks' prediction loss.
Despite its simplicity, we show several interesting phenomena by varying sample sizes and covariate shifts in this setting.
See Figure \ref{fig_intro_sample_size} for an illustration.
Consequently, using our precise loss estimates, we observe several qualitative properties of hard parameter sharing for varying datasets' properties.
%\begin{enumerate}

\textit{Sample efficiency (Example \ref{ex_same_cov})}:
	One advantage of combining multiple datasets is that the requirement for labeled data reduces compared to single-task learning, a phenomenon that \citet{ZSSGM18} has observed empirically.
	Our results further imply that HPS's sample efficiency depends on model-specific variances across tasks vs. the noise variance, and is generally high when the latter is large.

\textit{Sample size ratio (Example \ref{ex_sample_ratio})}: Increasing one task's sample size does not always help to reduce another task's loss. In a simplified setting, we find that the task loss either decreases first before increasing afterwards, or decreases monotonically depending on how fast the bias increases. These two trends result from different bias-variance tradeoffs. This result is surprising because previous generalization bounds in MTL typically scale down as the sample sizes of all tasks increase, thus do not apply for different sample sizes.

@@

% plus an identity matrix,  due to lack of freeness
%\HZ{add why this is challenging}

Finally, we discuss the practical implications of our work.
Our sample size ratio study implies a concrete progressive training procedure that gradually adds more data until performance drops.
For example, in the setting of Figure \ref{fig_intro_sample_size_b}, this procedure will stop right at the minimum of the local basin.
We conduct further studies of this procedure on six text classification datasets and observe that it reduces the computational cost by $65\%$ compared to a standard round-robin training procedure while keeping the average accuracy of all tasks simultaneously.

\textbf{Proof techniques.}
There are two main ideas in our analysis. The proof of our first result uses a geometric intuition that hard parameter sharing finds a ``rank-$r$'' approximation of the datasets.
We carefully keep track of the concentration error between the global minimum of $f(A, B)$ and its population version.
The proof of our second result is significantly more involved because of different sample sizes and covariate shifts. Using recently developed techniques from random matrix theory \cite{Anisotropic}, we show that the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts converges to a deterministic diagonal matrix asymptotically (cf. Theorem \ref{thm_main_RMT}). % Moreover, we can obtain a sharp bound on the concentration error.
% to obtain a sharp estimate on the..., which is commonly referred to as the \emph{local law}.
%\HZ{add several sentences on the technical insight}
One limitation of our analysis is that in Example \ref{ex_sample_ratio}, there is an error term that can result in vacuous bounds equation for very small $n_1$ (cf. equation \eqref{cor_MTL_error})).
We believe that our analysis has provided important initial insight and it is an interesting question to tighten our result (see Section \ref{sec_conclude} for more discussion).
